Enzo: An Adaptive Mesh Refinement Code for Astrophysics

### 2013-06-25 tests-fld-ifront.tex
Finished up all of the test problems!

This test problem examines \enzo's flux-limited diffusion radiation
transport and associated ionization chemistry solvers on a
cosmological I-front test.  This corresponds to test problem 4.6 from
\cite{ReynoldsHayesPaschosNorman2009}, run using the cosmological
deceleration parameter $q_0 = 0.05$ and initial redshift $z_i=10$.
Here the physics of interest is the expansion of an \ion{H}{2} region
in a uniform medium around a single monochromatic ionizing source
(with frequency $h\nu = 13.6$ eV).  The I-front propagates rapidly at
first, approaching 90\% of the Str{\" o}mgren radius by a scaled
redshift of $-\log\left[(1+z)/(1+z_i)\right] \approx 0.15$, before
cosmological expansion overtakes ionization, pushing the Str{\"
  o}mgren radius outward faster than the I-front can propagate.  The Str{\" o}mgren radius is analytically given by
\[
   r_S(t) = \left[\frac{3\dot{N}_{\gamma}}{4\pi \alpha_{\rm B}
   n_{\rm H}(t)^2}\right]^{1/3}, 
\]
where $\dot{N}_{\gamma} = 5 \times 10^{48}$ photon sec$^{-1}$ is the
strength of the ionizing source, $\alpha_{\rm B} = 2.52 \times 10^{-13}$
s$^{-1}$ is the \ion{H}{2} recombination rate, and $n_H(t)$ is the
proper number density of hydrogen.  We then define $\lambda =
\alpha_{\rm B}\, n_{H,i}\, /\, H_0\, / (1+z_i)$, where the subscript $i$
indicates quantities at the initial redshift.  The analytical
solution for the I-front position as a function of time is then

@@

The simulation parameters are: box size $L_i\approx 27$ kpc, Hubble
constant $H_0 = 100$ km s$^{-1}$ Mpc$^{-1}$, and cosmological constants $\Omega_m = 0.1$, $\Omega_\Lambda=0$ and $\Omega_b = 0.1$.  The
initial values for the simulation are a radiation energy density
$E_{r,i} = 10^{-35}$ erg cm$^{-3}$, temperature $T_i = 10^4$ K,
density $\rho_{b,i} = 2.35 \times 10^{-28}$ g cm$^{-3}$, and an
ionized hydrogen fraction $n_{\rm HII}/n_{\rm H} = 0$.

>>>>

This test problem examines \enzo's flux-limited diffusion radiation
transport and associated ionization chemistry solvers on a
cosmological I-front test.  This corresponds to test problem 4.6 from
\cite{ReynoldsHayesPaschosNorman2009}, run using the cosmological
deceleration parameter $q_0 = 0.05$ and initial redshift $z_i=10$.
In this test problem, the physics of interest is the expansion of an \ion{H}{2} region
in a uniform medium around a single monochromatic ionizing source
(with frequency $h\nu = 13.6$ eV).  The I-front propagates rapidly at
first, approaching 90\% of the Str{\" o}mgren radius by a scaled
redshift of $-\log\left[(1+z)/(1+z_i)\right] \approx 0.15$, before
cosmological expansion overtakes ionization, pushing the Str{\"
  o}mgren radius outward faster than the I-front can propagate.  The Str{\" o}mgren radius is analytically given by
\[
   r_S(t) = \left[\frac{3\dot{N}_{\gamma}}{4\pi \alpha_{\rm B}
   n_{\rm H}(t)^2}\right]^{1/3}, 
\]
where $\dot{N}_{\gamma} = 5 \times 10^{48}$ photon sec$^{-1}$ is the
strength of the ionizing source, $\alpha_{\rm B} = 2.52 \times 10^{-13}$
s$^{-1}$ is the \ion{H}{2} recombination rate, and $n_H(t)$ is the
proper number density of hydrogen.  We then define $\lambda =
\alpha_{\rm B}\, n_{H,i}\, /\, H_0\, / (1+z_i)$, where the subscript $i$
indicates quantities at the initial redshift.  The analytical
solution for the I-front position as a function of time is then

@@

The simulation parameters are: box size $L_i\approx 27$ kpc, Hubble
constant $H_0 = 100$ km s$^{-1}$ Mpc$^{-1}$, and cosmological parameters $\Omega_m = 0.1$, $\Omega_\Lambda=0$, and $\Omega_b = 0.1$.  The
initial values for the simulation are a radiation energy density
$E_{r,i} = 10^{-35}$ erg cm$^{-3}$, temperature $T_i = 10^4$ K,
density $\rho_{b,i} = 2.35 \times 10^{-28}$ g cm$^{-3}$, and an
ionized hydrogen fraction $n_{\rm HII}/n_{\rm H} = 0$.

### 2013-06-25 tests-onezonecollapse.tex
Finished up all of the test problems!

The one-zone collapse test simulates the collapse of a
self-gravitating gas cloud using a semi-analytic model for the
evolution of gas density and adiabatic heat input as a function of
time.  It is designed to test the chemistry and cooling modules over a
wide range in densities and over physically-motivated timescales.
Because this test disables the hydrodynamic and gravity solvers and
uses a simple model for the density evolution, it is far faster than
running a true collapse simulation.  The density evolution is based on
the self-similar Larson-Penston \citep{1969MNRAS.145..271L,
1969MNRAS.144..425P} solution for isothermal collapse with a
modification to account for the efficiency with which the heat
introduced by compression can be radiated away
\citep{1983ApJ...265.1047Y}.  Our implementation, described briefly
here, follows the work of \citet{2005ApJ...626..627O}, and we direct
the interested reader to this paper for further details.  The
evolution of the gas density, $\rho$, is given by

@@

\begin{equation} \label{eqn.tcol}
t_{\rm col} = \frac{t_{\rm dyn}}{\sqrt{1 - f}},
\end{equation}
In this equation, $t_{\rm dyn}$ is the dynamical time and is expressed as
\begin{equation}
t_{\rm dyn} = \sqrt{\frac{3 \pi}{32 G \rho}}.
\end{equation}

@@

\begin{figure}
  \begin{center}
    \includegraphics[width=1.0\textwidth]{figures/OneZoneCollapseTest.eps}
  \end{center}
  \caption{Evolution of temperature versus number density for a
one-zone collapse test for gas at metallicities from 0 to 10$^{-2}
Z_{\odot}$.  This test is based on the results of
\citet{2005ApJ...626..627O}, and approximates the collapse of a
self-gravitating, cooling gas cloud.  This test problem uses \enzo's
primordial chemistry network with tabulated metal cooling rates
calculated with the \texttt{Cloudy} code, and compares favorably to
the results of \citet{2005ApJ...626..627O}.}
  \label{fig.onezone}
\end{figure}

>>>>

The one-zone collapse test simulates the collapse of a
self-gravitating gas cloud using a semi-analytic model for the
evolution of gas density and adiabatic heat input as a function of
time.  It is designed to test the chemistry and cooling modules over a
wide range in densities and over physically-motivated timescales.
Because this test disables the hydrodynamic and gravity solvers and
uses a simple model for the density evolution, it is far faster than
running a true collapse simulation.  The density evolution is based on
the self-similar Larson-Penston solution for isothermal collapse \citep{1969MNRAS.145..271L,
1969MNRAS.144..425P} with a
modification to account for the efficiency with which the heat
introduced by compression can be radiated away
\citep{1983ApJ...265.1047Y}.  Our implementation, described briefly
here, follows the work of \citet{2005ApJ...626..627O}, and we direct
the interested reader to this paper for further details.  The
evolution of the gas density, $\rho$, is given by

@@

\begin{equation} \label{eqn.tcol}
t_{\rm col} = \frac{t_{\rm dyn}}{\sqrt{1 - f}},
\end{equation}
In this equation, $t_{\rm dyn}$ is the dynamical time for the collapse
of a spherical cloud, and is expressed as
\begin{equation}
t_{\rm dyn} = \sqrt{\frac{3 \pi}{32 G \rho}}.
\end{equation}

@@

\begin{figure}
  \begin{center}
    \includegraphics[width=0.8\textwidth]{figures/OneZoneCollapseTest.eps}
  \end{center}
  \caption{Evolution of temperature versus number density for a
one-zone collapse test for gas at metallicities from 0 to 10$^{-2}
Z_{\odot}$.  This test is based on the results of
\citet{2005ApJ...626..627O}, and approximates the collapse of a
self-gravitating, cooling gas cloud.  This test problem uses \enzo's
primordial chemistry network with tabulated metal cooling rates
calculated with the \texttt{Cloudy} code, and compares favorably to
the results of \citet{2005ApJ...626..627O}.}
  \label{fig.onezone}
\end{figure}

### 2013-06-25 tests-orszagtang.tex
Finished up all of the test problems!

\subsubsection{MHD: Orszag-Tang Vortex}
\label{sec.tests.mhd}
Figure~\ref{fig.orszag} shows the Orszag-Tang vortex problem, a
classic MHD test problem described by \citet{Orszag79}.  The left panel shows
the result using \enzo's constrained transport MHD method, while the right panel shows the result using
\enzo's implementation of Dedner MHD.
The Orszag-Tang vortex test shows that significant small scale structure can be generated in MHD
from large scale initial perturbations, and is used to compare the
effective resolution of different MHD schemes.  The test begins with uniform
density, $\rho_0=25/(36\pi)$, and pressure, $P_0=5/(12\pi)$ (as with
other tests in this section, in the absence of gravity or chemistry we use
arbitrary units).  There is a
single rotational mode in the velocity, and two in the magnetic field:
${\bf v}_0 = (-\sin(2\pi y) \hat{x},\, \sin(2\pi x) \hat{y})$,
${\bf B}_0 = (-\sin(2\pi y) \hat{x},\, \sin(4\pi x) \hat{y})$.
The simulation is evolved to $t=0.48$ (in arbitrary units).  One can see that the structures
are accurately represented as compared to, for example,
\citet{Toth00}, and that the resolution of shocks is comparable in
both methods.

>>>>

\subsubsection{MHD: Orszag-Tang Vortex}
\label{sec.tests.mhd}
Figure~\ref{fig.orszag} shows the Orszag-Tang vortex problem \citep{Orszag79}.  The left panel shows
the result using \enzo's constrained transport MHD method, while the right panel shows the result using
\enzo's implementation of Dedner MHD.
The Orszag-Tang vortex test is a classig MHD test problem, and shows that significant small scale structure can be generated in MHD
from large scale initial perturbations.  It is often used to compare the
effective resolution of different MHD schemes.  The test begins with uniform
density, $\rho_0=25/(36\pi)$, and pressure, $P_0=5/(12\pi)$ (as with
other tests in this section, in the absence of gravity or chemistry/cooling we use
arbitrary units).  There is a
single rotational mode in the velocity, and two in the magnetic field:
${\bf v}_0 = (-\sin(2\pi y) \hat{x},\, \sin(2\pi x) \hat{y})$,
${\bf B}_0 = (-\sin(2\pi y) \hat{x},\, \sin(4\pi x) \hat{y})$.
The simulation is evolved to $t=0.48$ (in arbitrary units).  One can see that the structures
are accurately represented as compared to, for example,
\citet{Toth00}, and that the resolution of shocks is comparable in
both methods.

### 2013-06-25 tests-photoevaporation.tex
Finished up all of the test problems!

The photo-evaporation of dense clumps of gas is prevalent in radiation
hydrodynamics simulations, and this test problem examines the
ionization front propagation into a dense clump, shadowing effects
behind the clump, and the hydrodynamic response on the clump from
photo-heating, all using \enzo's Moray ray-tracing module.  The
problem setup is the same as Test 7 in the Cosmological Radiative
Transfer Comparison Project \citep{IlievEtAl2009} and
\citet{Wise11_Moray}.  The simulation domain is 6.6~kpc on a side with
an ambient medium of pure neutral hydrogen of density $n_{\rm H} =
2 \times 10^{-4}\; \cubecm$ and temperature $T = 8000 \unit{K}$.  We
place a spherical overdensity in hydrostatic equilibrium with the
ambient medium.  It has a radius $r = 0.8 \unit{kpc}$, hydrogen
density $n_{\rm H} = 0.04\ \cubecm$, and temperature $T = 40$ K, and
is centered at $(x,y,z) = (5, 3.3, 3.3) \unit{kpc}$.
In \citet{IlievEtAl2009} all of the codes used a fixed $128^3$ grid to
ease the comparison, but in this test to demonstrate a higher
resolution AMR solution, we employ a $128^3$ grid with two additional
levels of refinement for cells with a baryon mass greater than 1.5
(method 2 in Section~\ref{sec:refinement_criteria}). This test is run
for 15 Myr.

This cloud is subject to radiation from a point source at the center
of the $x=0$ boundary with an ionizing photon luminosity
$\dot{N}_\gamma = 3 \times 10^{51}$ photons s$^{-1}$, corresponding to
a flux $F_0 = 10^6 \unit{photons s}^{-1} \unit{cm}^{-2}$ at the clump
surface closest to the radiation source.  The radiation source has a
spectrum of a $T = 10^5 \unit{K}$ blackbody, and we use four energy
groups with the following mean energies and relative luminosities:
$E_i = (17.98, 31.15, 49.09, 76.98) \unit{eV}, L_i/L = (0.23, 0.36,
0.24, 0.06)$ that are optimized to reduce errors in the solutions with
a full spectrum and energy discretization \citep{Mirocha12}.
(Note that this choice of energy groups is different from Wise \& Abel 2011.) \nocite{Wise11_Moray}  We use a minimum angular resolution of 10 rays
per cell and a constant radiative transfer timestep of 25 kyr.  Figure
\ref{fig:shadowing} depicts the clump at $t = 15$ Myr with the
outer layers expanding after being photo-heated.  It also shows the
sharp shadowing effects of the dense clump in the neutral fraction
plot that is representative of ray tracing techniques.

>>>>

The photo-evaporation of dense clumps of gas is prevalent in radiation
hydrodynamics simulations, and this test problem examines the
ionization front propagation into a dense clump, shadowing effects
behind the clump, and the hydrodynamic response on the clump from
photo-heating, all using \enzo's Moray ray-tracing module.  The
problem setup is the same as Test 7 in the Cosmological Radiative
Transfer Comparison Project \citep{IlievEtAl2009} and
\citet{Wise11_Moray}.  The simulation domain is 6.6~kpc on a side with
an ambient medium of pure neutral hydrogen of density $n_{\rm H} =
2 \times 10^{-4}\; \cubecm$ and temperature $T = 8000 \unit{K}$.  We
place a spherical overdensity in hydrostatic equilibrium with the
ambient medium.  It has a radius $r = 0.8 \unit{kpc}$, hydrogen
density $n_{\rm H} = 0.04\ \cubecm$ (i.e., overdensity of 200), and temperature $T = 40$ K, and
is centered at $(x,y,z) = (5, 3.3, 3.3) \unit{kpc}$.
In \citet{IlievEtAl2009} all of the codes used a fixed $128^3$ grid to
ease the comparison, but in this test to demonstrate a higher
resolution AMR solution, we employ a $128^3$ grid with two additional
levels of refinement by factors of 2 for cells with a baryon mass greater than 1.5
(method 2 in Section~\ref{sec:refinement_criteria}). This test is run
for 15 Myr.

The cloud is subject to radiation from a point source at the center
of the $x=0$ boundary with an ionizing photon luminosity
$\dot{N}_\gamma = 3 \times 10^{51}$ photons s$^{-1}$, corresponding to
a flux $F_0 = 10^6 \unit{photons s}^{-1} \unit{cm}^{-2}$ at the clump
surface closest to the radiation source.  The radiation source has a
spectrum of a $T = 10^5 \unit{K}$ blackbody, and we use four energy
groups with the following mean energies and relative luminosities:
$E_i = (17.98, 31.15, 49.09, 76.98) \unit{eV}, L_i/L = (0.23, 0.36,
0.24, 0.06)$ that are optimized to reduce errors in the solutions with
a full spectrum and energy discretization \citep{Mirocha12}.
(Note that this choice of energy groups is different from those used
in Wise \& Abel 2011.) \nocite{Wise11_Moray}  We use a minimum angular resolution of 10 rays
per cell and a constant radiative transfer timestep of 25 kyr.  Figure
\ref{fig:shadowing} depicts the clump at $t = 15$ Myr with the
outer layers expanding after being photo-heated.  It also shows the
sharp shadowing effects of the dense clump in the neutral fraction
plot that is representative of ray tracing techniques.

### 2013-06-25 tests-zeldovichpancake.tex
Finished up all of the test problems!

\subsubsection{Zel`dovich Pancake}
\label{sec.tests.pancake}

The Zel`dovich pancake \citep{1970A&A.....5...84Z} is particularly
relevant to cosmological simulations, because it includes many
features that are critical to structure formation: hydrodynamics,
expansion, and self gravity.  This problem represents the formation of an ideal,
isolated caustic, and is thus a useful proxy for much more
complicated structures in full 3-dimensional cosmological simulations,
such as the collapse of gas onto a cosmological halo or filament.

@@

In the equations above, $\rho_0$ is the background density, z$_c$ is a
free parameter and is the redshift where the sheet forms a caustic
(i.e., where it `pancakes'), z is the redshift of initialization, x$_l$ is the
Lagrangian mass coordinate, $k = 2 \pi / \lambda$ (where $\lambda$ is
the perturbation wavelength), and H$_0$ is the value of the Hubble
constant at $z = 0$.  Note that this solution is expressed in terms of 
Lagrangian positions, so one needs to convert this into the Eulerian
coordinates, x$_e$, that are more useful to \enzo:
\begin{equation}
x_e = x_l - \frac{1 + z_c}{1 + z} \frac{\sin(k x_l)}{k}
\end{equation}

We note that the solution described above is exact up to the point of
caustic formation.  In Figure~\ref{fig.pancake}, we show the results
of a test of the adaptive mesh version of \enzo's Zel'dovich Pancake test.  A
one-dimensional box of length $64$~Mpc/h is initialized at $z = 20$ in
an $\Omega_M = 1$ universe with $h = 0.5$ and a background temperature
of 100~K, $\rho_0 = \rho_c$.  The simulation is initialized with 64
grid cells, refining by factors of four using criteria based on cell
mass and the presence of shocks, for a maximum of 2 levels (i.e., an
equivalent maximum resolution of 1024 grid cells).  The simulation is
evolved to z$ = 0$ using the PPM hydro method.

>>>>

\subsubsection{Zel'dovich Pancake}
\label{sec.tests.pancake}

The Zel'dovich pancake \citep{1970A&A.....5...84Z} is particularly
relevant to cosmological simulations, because it includes many
features that are critical to structure formation: hydrodynamics,
expansion, and self gravity.  This problem represents the formation of an ideal,
isolated caustic, and is thus a useful proxy for much more
complicated structures in full 3-dimensional cosmological simulations,
such as the collapse of gas onto a cosmological halo or filament.

@@

In the equations above, $\rho_0$ is the background density, z$_c$ is a
free parameter and is the redshift where the sheet forms a caustic
(i.e., where it `pancakes'), z is the redshift of initialization, x$_l$ is the
Lagrangian mass coordinate, $k = 2 \pi / \lambda$ (where $\lambda$ is
the perturbation wavelength), and H$_0$ is the value of the Hubble
constant at $z = 0$.  Note that this solution is expressed in terms of 
Lagrangian positions, so one needs to convert this into the Eulerian
coordinates, x$_e$, that are more useful to a grid-based calculation:
\begin{equation}
x_e = x_l - \frac{1 + z_c}{1 + z} \frac{\sin(k x_l)}{k}
\end{equation}

We note that the solution described above is exact up to the point of
caustic formation.  In Figure~\ref{fig.pancake}, we show the results
of a test of the adaptive mesh version of \enzo's Zel'dovich Pancake test.  A
one-dimensional box of length $64$~Mpc/h is initialized at $z = 20$ in
an $\Omega_M = 1$ universe with $h = 0.5$ and a background temperature
of 100~K, with a background density of $\rho_0 = \rho_c$.  The simulation is initialized with 64
grid cells, refining by factors of four using criteria based on cell
mass and the presence of shocks, for a maximum of 2 levels (i.e., an
equivalent maximum resolution of 1024 grid cells).  The simulation is
evolved to z$ = 0$ using the PPM hydro method.
